Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-07-30 doc/study/study.tex
Another good source regarding hashing algorithms.
Integrated Sepps comments.

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent look-up of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased. (The actual
		accommodation of E-Mails -- another big problem with horizontal scaling
		-- can not be considered here.)

@@

		\subsection{Criteria}
		\label{sec:evaluation:criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution. The criteria will be classified into five
			categories (Data model, features, security \& robustness, user
			interface and miscellaneous) and numbered for easier reference.

			There is some meta-data like website, programming language, last
			update in VCS and number of commits, which is not numbered and
			compared. The used programming language would be a neutral
			attribute but gets a criterion, because before the development of
			\emph{Møil}, a decision was made against deploying further PHP
			applications to accomplish a more maintainable and secure web
			hosting environment\footnote{\urlPHPBashing}.

			\subsubsection{Data model (1)}
				From a database engineering perspective, it is in general
				favourable to design a database with a \crit{normalized
				relational database schema}{1.1} \cite{dbnorm}. For a query to
				check the existence of a mailbox for example, a JOIN or nested
				query would be necessary, which has a very small performance
				impact. However, the schema normalization prevents data
				redundancies and inconsistencies. \emph{postfix.admin} does not
				have a normalized schema, so any solution which has one can not
				be \crit{compatible to \emph{postfix.admin}}{1.2}. A
				compatibility could be useful, if there is a big database
				already administered with \emph{postfix.admin}, which shall be
				migrated to another tool. Another possibility for this is an
				\crit{import of \emph{postfix.admin} data}{1.3}. When it comes
				to updates to the administration interface, \crit{means to
				handle database schema updates}{1.4} ease the update process
				(and therefore increase the probability of the actual
				conduction of an update). Last but not least, \crit{systematic
				database independence}{1.5} abstracts from raw \ac{SQL} queries
				and allows a free choice of the \ac{DBMS} and integration with
				other software.

			\subsubsection{Features (2)}
				There are features, without which a mail server administration
				user interface would not work at all and yet I list them,
				because it is more about analyzing, how these are implemented.
				The \crit{basic domain, mailbox and alias management}{2.1} is
				one of them. The \crit{password changing ability for
				users}{2.2} is a feature, which is also quite basic and the
				first one legitimating letting users access the administration
				front-end. Some solutions also integrate into popular
				self-hosted web mailer software for password changing. When
				users are allowed to access the administration front-end,
				\crit{access control}{2.3} or authorization is needed to
				selectively restrict the access to resources. It could also be
				possible to delegate administrative access on domains or
				resources in general to certain users.
				\s

@@

			\subsubsection{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \texttt{MD5} and \texttt{SHA1} based methods are
				considered unsecure, \texttt{SHA256}, \texttt{SHA384},
				\texttt{SHA512} and \texttt{bcrypt} are
				recommended\footnote{\url{http://www.keylength.com/en/8/}}.
				\crit{Test coverage}{3.2} is important for this evaluation and
				for security purposes, because it lowers the probability of
				bugs per lines of code by trend. As \crit{Containability}{3.3},
				the modular independence (e.g. independence of the web
				interface from the Message Store file system) from a security
				perspective is perceived.

@@

				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterion; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone, was a little sloppy.

@@

				Further information on \emph{Møil} is to be found under
				\textref{sec:moeil}.

@@

				PostVis Admin is the best acknowledgement for the non-PHP
				policy it was decided on when searching for a
				\emph{postfix.admin} replacement. The code is very bad
				structured and full of redundancies. There is no \crit{schema
				normalization}{1.1} and the schema seems \crit{compatible to
				\emph{postfix.admin}}{1.2} but is likely to depart from it in
				some details. \ac{SQL} injections are findable just by checking
				for criteria like if there is an \crit{access control}{2.3}
				feature, which is basically existing but not very effective,
				because it is not implemented securely. (Any logged in user
				(also non-superadmin users) seem to be able to delete any other
				user.) The default \crit{password hashing}{3.1} algorithm is
				\texttt{md5-crypt} and there are features, which do not work
				without shell execution privileges on the mail system servers
				(e.g. \texttt{amavisd-release}, so the solution is not really
				\crit{containable}{3.3}.

@@

	\section{Møil}
	\label{sec:moeil}
		The development of \emph{Møil} has been begun in May, 2013 as an Open
		Source project licensed under the terms of the AGPL, Version 3
		\cite{agpl}.

		% Rails

		It is written in Ruby\footnote{\url{https://www.ruby-lang.org}}
		utilizing the popular \ac{Rails} web framework, which has some features
		that emphasize it against other frameworks or web development
		approaches. It promotes the development by \acs{REST} \cite{rest}
		criteria, supports test driven development and implements many
		well-known software engineering design patterns and methods of agile
		software development as Active Record, \ac{CoC}, \ac{DRY} and \ac{MVC}.
		Additionally, the \ac{Rails} community is quite vivid and there is an
		huge ecosystem of extensions for many recurrent problems.
		\s

@@

		\subsection{Features}
		\label{sec:moeil:features}
			At the time of writing of this document, \emph{Møil} comes with the
			below-mentioned features.

@@

			The code is strictly modularized according to the \ac{MVC} pattern,
			like it is encouraged by \ac{Rails}.

			The model connects to the \ac{DBMS} via TCP/IP or Unix Sockets. The
			abstractly noted queries are translated to the respective \ac{SQL}
			(or NoSQL) dialect. All communication with external services like
			\emph{postfix}, \emph{dovecot} or the relocation daemon (see
			\textref{sec:moeil:model:relocation}) happen solely with the
			\ac{DBMS} as an intermediary.

			The view templates are written in
			Haml\footnote{\url{http://haml.info}}, styling is done in SASS and
			CoffeeScript is used for client-side dynamic functionality. Markup,
			styling and scripts are delivered to the client after being
			processed by the \ac{Rails} asset pipeline, which converts HAML to
			HTML, SASS to CSS and CoffeeScript to JavaScript. It also
			compresses the output and packs it in as few as possible files to
			keep the requests to the server low.

			Controllers and their actions are routed to by URL path patterns
			and are responsible for updating the models' state and providing
			the according data for the views.

			The search is either solved through using the \ac{DBMS} via
			\ac{SQL} or by using \emph{Elasticsearch}, which is an indexed
			search server based on \emph{Apache Lucene}.

		\subsection{Data model}
			% Data model
			%	Structure / ERD

@@

			The model classes will be examined individually in the following
			subsections.
			\s

			In this chapter, whenever a resource type is mentioned, which
			corresponds to an actual \ac{Rails} model class, it is written in a
			monospace font and -- as the class name -- capitalized, like
			\texttt{Mailbox}. Attributes of models are also in monospace font,
			but lowercase, as the \texttt{username} attribute of a
			\texttt{Mailbox} for example.

@@

			\subsubsection{Mailboxes}
				A \texttt{Mailbox} is connected to its \texttt{Domain} and
				possibly also to one \texttt{Relocation} and has the attributes
				\texttt{username} (which is the part in front of the
				domain in the E-Mail address), \texttt{encrypted\_password}
				(containing the password hash), \texttt{name} (for a given name
				for example), \texttt{mail\_location} (which is empty by
				default but can be used to move the users mail data location),
				\texttt{quota}, \texttt{active}, \texttt{admin} and timestamp
				attributes for saving the date of creation or the last update.

@@

			\subsubsection{Versions}
				As seen in the graph, \texttt{Domains}, \texttt{Mailboxes},
				\texttt{Aliases} and \texttt{Permissions} are versioned, which
				means that changes to resources of any of these types are
				tracked in the audit log and are revertable for administrators.
				On any change, a reference to the corresponding resource
				(\texttt{item\_id} and \texttt{item\_type}), the type of change
				(\texttt{event}), the transacting individual
				(\texttt{whodunnit}), the serialized \texttt{object} (only on
				deletions), the \texttt{object\_changes} (only on updates) and
				the point in time (\texttt{created\_at}) are saved.

@@

			%	Migrations
			%	sha512-crypt

@@

			\subsubsection{Relevant configs for \emph{postfix} and \emph{dovecot}}
				The relevant configuration files for \emph{postfix} and
				\emph{dovecot} (especially regarding to the \ac{SQL} statements
				compatible to \emph{Møil}) can be found in the source code
				repository on GitHub in the folders
				\texttt{doc/postfix}\footnote{\url{https://github.com/nning/moeil/tree/master/doc/postfix}}
				and
				\texttt{doc/dovecot}\footnote{\url{https://github.com/nning/moeil/tree/master/doc/dovecot}}.

>>>>

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent look-up of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased. The actual
		accommodation of E-Mails -- another big problem with horizontal scaling
		-- can not be considered here.

@@

		\subsection{Criteria}
		\label{sec:evaluation:criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution. The criteria will be classified into four
			categories (Data model, features, security \& robustness and user
			interface) and numbered for easier reference. An example for the
			textual distinguation is the criterion \crit{normalized database
			schema}{1.1} as it is written here. There is some meta-data like
			website, programming language, last update in VCS and number of
			commits, which is not numbered and compared. The used programming
			language would be a neutral attribute but gets a criterion, because
			before the development of \emph{Møil}, a decision was made against
			deploying further PHP applications to accomplish a more
			maintainable and secure web hosting
			environment\footnote{\urlPHPBashing}.

			\subsubsection{Data model (1)}
				From a database engineering perspective, it is in general
				favourable to design a data\-base with a \crit{normalized
				relational database schema}{1.1} \cite{dbnorm}. For a query to
				check the existence of a mailbox for example, a JOIN or nested
				query would be necessary, which has a very small performance
				impact. However, the schema normalization prevents data
				redundancies and inconsistencies. \emph{postfix.admin} does not
				have a normalized schema, so any solution which has one can not
				be \crit{compatible to \emph{postfix.admin}}{1.2}. A
				compatibility could be useful, if there is a big database
				already administered with \emph{postfix.admin}, which shall be
				migrated to another tool. Another possibility for this is an
				\crit{import of \emph{postfix.admin} data}{1.3}. When it comes
				to updates to the administration interface, \crit{means to
				handle database schema updates}{1.4} ease the update process
				(and therefore increase the probability of the actual
				conduction of an update). Last but not least, \crit{systematic
				database independence}{1.5} abstracts from raw \ac{SQL} queries
				and allows a free choice of the \ac{DBMS} and integration with
				other software.

			\subsubsection{Features (2)}
				There are features, which no mail server administration user
				interface could spare and yet they will be listed here, because
				it is important to analyze, how these are implemented and how
				the implementations differ by solution. The \crit{basic
				domain, mailbox and alias management}{2.1} is one of them. The
				\crit{password changing ability for users}{2.2} is a feature,
				which is also quite basic and the first one legitimating
				letting users access the administration front-end. Some
				solutions also integrate into popular self-hosted web mailer
				software for password changing. When users are allowed to
				access the administration front-end, \crit{access control}{2.3}
				or authorization is needed to selectively restrict the access
				to resources. It could also be possible to delegate
				administrative access on domains or resources in general to
				certain users.

@@

			\subsubsection{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \texttt{MD5} and \texttt{SHA1} based methods are
				considered unsecure, \texttt{SHA256}, \texttt{SHA384},
				\texttt{SHA512} and \texttt{bcrypt} are
				recommended\footnote{\url{http://www.keylength.com/en/8/}}\textsuperscript{,}\footnote{\url{http://codahale.com/how-to-safely-store-a-password}}.
				\crit{Test coverage}{3.2} is important for this evaluation and
				for security purposes, because it lowers the probability of
				bugs per lines of code by trend. As \crit{Containability}{3.3},
				the modular independence (e.g. independence of the web
				interface from the Message Store file system) from a security
				perspective is perceived.

@@

				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterion; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone is not fully
				elaborated and has some minor flaws.

@@

				Further information on \emph{Møil} is to be found in the
				section \textref{sec:moeil}.

@@

				\emph{PostVis Admin} is the best acknowledgement for the
				non-PHP policy it was decided on when searching for a
				\emph{postfix.admin} replacement. The code is very bad
				structured and full of redundancies. There is no \crit{schema
				normalization}{1.1} and the schema seems \crit{compatible to
				\emph{postfix.admin}}{1.2} but is likely to depart from it in
				some details. \ac{SQL} injections are findable just by checking
				for criteria like if there is an \crit{access control}{2.3}
				feature, which is basically existing but not very effective,
				because it is not implemented securely. (Any logged in user
				(also non-superadmin users) seem to be able to delete any other
				user.) The default \crit{password hashing}{3.1} algorithm is
				\texttt{md5-crypt} and there are features, which do not work
				without shell execution privileges on the mail system servers
				(e.g. \texttt{amavisd-release}, so the solution is not really
				\crit{containable}{3.3}.

@@

	\section{Møil}
	\label{sec:moeil}
		The development of \emph{Møil} has been begun in May, 2013 as an Open
		Source project licensed under the terms of the AGPL, Version 3
		\cite{agpl}. It is written in
		Ruby\footnote{\url{https://www.ruby-lang.org}} utilizing the popular
		\ac{Rails} web framework, which has some features that emphasize it
		against other frameworks or web development approaches. It promotes the
		development by \acs{REST} \cite{rest} criteria, supports test driven
		development and implements many well-known software engineering design
		patterns and methods of agile software development as Active Record,
		\ac{CoC}, \ac{DRY} and \ac{MVC}. Additionally, the \ac{Rails}
		community is quite vivid and there is an huge ecosystem of extensions
		for many recurrent problems.

@@

		\subsection{Features}
		\label{sec:moeil:features}
			At the time of writing of this document, \emph{Møil} comes with the
			below-mentioned features:

@@

			The code is strictly modularized according to the \ac{MVC} pattern,
			like it is encouraged by \ac{Rails}. The model connects to the
			\ac{DBMS} via TCP/IP or Unix Sockets. The abstractly noted queries
			are translated to the respective \ac{SQL} (or NoSQL) dialect. All
			communication with external services like \emph{postfix},
			\emph{dovecot} or the relocation daemon (see
			\textref{sec:moeil:model:relocation}) happen solely with the
			\ac{DBMS} as an intermediary. The view templates are written in
			Haml\footnote{\url{http://haml.info}}, styling is done in SASS and
			CoffeeScript is used for client-side dynamic functionality. Markup,
			styling and scripts are delivered to the client after being
			processed by the \ac{Rails} asset pipeline, which converts HAML to
			HTML, SASS to CSS and CoffeeScript to JavaScript. It also
			compresses the output and packs it in as few as possible files to
			keep the requests to the server low. Controllers and their actions
			are routed to by URL path patterns and are responsible for updating
			the models' state and providing the according data for the views.
			The search is either solved through using the \ac{DBMS} via
			\ac{SQL} or by using \emph{Elasticsearch}, which is an indexed
			search server based on \emph{Apache Lucene}.

		\subsection{Data model}
			% Data model
			%	Structure / ERD
			%	Migrations
			%	sha512-crypt

@@

			\noindent
			The model classes will be examined individually in the following
			subsections. In this chapter, whenever a resource type is
			mentioned, which corresponds to an actual \ac{Rails} model class,
			it is written in a monospace font and -- as the class name --
			capitalized, like \texttt{Mailbox}. Attributes of models are also
			in monospace font, but lowercase, as the \texttt{username}
			attribute of a \texttt{Mailbox} for example.

@@

			\subsubsection{Mailboxes}
				A \texttt{Mailbox} is connected to its \texttt{Domain} and
				possibly also to one \texttt{Relocation} and has the attributes
				\texttt{username} (which is the part in front of the
				domain in the E-Mail address), \texttt{encrypted\_password}
				(containing the password hash), \texttt{name} (for a given name
				for example), \texttt{mail\_location} (which is empty by
				default but can be used to move the users mail data location),
				\texttt{quota}, \texttt{active}, \texttt{admin} and timestamp
				attributes for saving the date of creation or the last update.
				As a password hashing algorithm, currently only
				\texttt{sha512-crypt} is supported.

@@

			\subsubsection{Versions}
				As seen in the graph, \texttt{Domains}, \texttt{Mailboxes},
				\texttt{Aliases} and \texttt{Permissions} are versioned, which
				means that changes to resources of any of these types are
				tracked in the audit log and are revertable for administrators.
				On any change, a reference to the corresponding resource
				(\texttt{item\_id} and \texttt{item\_type}), the type of change
				(\texttt{event}), the transacting individual
				(\texttt{whodunnit}), the serialized \texttt{object} (only on
				deletions, can be used for undo), the \texttt{object\_changes}
				(only on updates) and the point in time (\texttt{created\_at})
				are saved.

@@

@@

			\subsubsection{Relevant configs for \emph{postfix} and \emph{dovecot}}
				The relevant configuration files for \emph{postfix} and
				\emph{dovecot} (especially regarding to the \ac{SQL} statements
				compatible to \emph{Møil}) can be found in the source code
				repository on GitHub in the folders \texttt{doc/postfix} and
				\texttt{doc/dovecot}\footnote{\url{https://github.com/nning/moeil/tree/master/doc}}.
